### Question 0

*  Draw a concepts diagram that uses all the following Github terminology 
   * Include any missing keywords that will simplify the concepts diagrams 


`Push`, `Repository`, `Clone`, `Pull`, `Pull Request`, `Branch`, `Merging`, `Github Client`, `README file`, `Private or Public`

[GitHub Concept Diagram](https://docs.google.com/presentation/d/15GAjEJdCxWzjh8ICHS-V4zBd09-5mCXu6zmlgOgTDPE/edit?usp=sharing)


#### Question 1

`matrix(c(1,2,3,4,5,6) , nrow = 3)`

* Running the expression produces the following matrix

|  |  |
|---|---|
| 1 | 4 |
| 2 | 5 |
| 3 | 6 |

* How can you modify the call to `matrix()` to produce the following matrix instead?

|   | col_1 |  col_2 |
| --- |---|---|
| row_1 | 1 | 2 |
| row_2 | 3 | 4 |
| row_3 | 5 | 6 |

* Note that you need to name the columns (col_1 and col_2) and name the rows (row_1, row_2, row_3)

Hint: Use the `?` symbol to invoke the matrix documentation

```{r}
# create matrix, sort values row-wise
mod_mat <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, byrow = TRUE)

# add labels
colnames(mod_mat) <- c("col_1", "col_2")
rownames(mod_mat) <- c("row_1", "row_2", "row_3")

# print
mod_mat
```


#### Question 2

* Load then sort the air quality data frame on its `Temp` and `Solar.R` columns in reverse order (largest to smallest values)
  * The function to sort a data frame is called order
* Display only the first 15 lines of your table


```{r}
# load data from set airquality
data("airquality")

# extract index values for columns `Temp` and `Solar.R` & sort into decreasing order
sort_temp_dec <- airquality$Temp[order(-airquality$Temp)]
sort_solar.r_dec <- airquality$Solar.R[order(-airquality$Solar.R)]

# create new matrix of ordered values in airquality
df_sort_temp_solar.r <- data.frame(airquality$Ozone, sort_solar.r_dec, airquality$Wind, sort_temp_dec, airquality$Month, airquality$Day)

# re-name columns 
names(df_sort_temp_solar.r)[1] <- "Ozone"
names(df_sort_temp_solar.r)[2] <- "Dec. Solar.R"
names(df_sort_temp_solar.r)[3] <- "Wind"
names(df_sort_temp_solar.r)[4] <- "Dec. Temp"
names(df_sort_temp_solar.r)[5] <- "Month"
names(df_sort_temp_solar.r)[6] <- "Day"

# print first 15 lines
head(df_sort_temp_solar.r, n = 15)
```


#### Question 3

* Sort the airquality data frame on its `Temp` in decreasing order and `Solar.R` in increasing order
* Display only the first 15 lines of your table

```{r}
# extract index values for column `Solar.R` & sort into increasing order
sort_solar.r_inc <- airquality$Solar.R[order(airquality$Solar.R)]

# create new matrix of ordered values in airquality
df_sort_temp_solar.r_inc <- data.frame(airquality$Ozone, sort_solar.r_inc, airquality$Wind, sort_temp_dec, airquality$Month, airquality$Day)

# re-name columns 
names(df_sort_temp_solar.r_inc)[1] <- "Ozone"
names(df_sort_temp_solar.r_inc)[2] <- "Inc. Solar.R"
names(df_sort_temp_solar.r_inc)[3] <- "Wind"
names(df_sort_temp_solar.r_inc)[4] <- "Dec. Temp"
names(df_sort_temp_solar.r_inc)[5] <- "Month"
names(df_sort_temp_solar.r_inc)[6] <- "Day"

# print first 15 lines
head(df_sort_temp_solar.r_inc, n = 15)
```


#### Question 4

* There are various ways to select a subset of observations from a data frame.

* Consult your [R Reference Card](https://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf), see `Data Selection and Manipulation` section.
  * What operations can you use to select all observations where the temperature is 72. Give at least two different answers to this question

(1) Boolean Scanning Method 
```{r}
# create boolean expression to index Temp data that contains 72 degrees, print result
temp72_a <- airquality[airquality$Temp == 72, ]

# print
temp72_a
```


(2) Using Function is.element()
```{r}
# using function to sort through Temp data column
temp72_b <- airquality[is.element(airquality$Temp, c('72')), ]

#print
temp72_b
```

#### Question 6

* You may have noticed when working with the `airqulity` data that some values show as `NA`
 * `NA` stands for not available, or missing values.
* A major part of data wrangling consists of cleaning missing values by either:
  * Dropping the lines that have missing values
   * Sometimes we can drop the column with missing values if the column is made of predominantly missing values
  * Imputing the missing values, which uses educated guesses (or more complex algorithms) to fill the missing values
  
* Find and remove all rows that are missing values for the `Solar.R` or `Ozone` variables
* Save the cleaned data to a new data frame called airquality_no_na
  * How many lines have been removed?

```{r}
# remove rows containing NA using na.omit() function and boolean statement
airquality_no_na <- na.omit(
  airquality[airquality$Solar.R != 'NA' & airquality$Ozone != 'NA', ]
  )

# print new data frame
head(airquality_no_na, n = 10)

# compute difference in dataframe row length
lines_removed <- nrow(airquality) - nrow(airquality_no_na)

# print number of lines removed
lines_removed
```


#### Question 7
* Let's use a different strategy and impute the missing value.
  * replace the missing values for Solar.R using that month's average.
  * Example:
    * The missing value for line 6 should be replaced with the average for month 5.
    * The missing value for line 97 should be replaced with the average for month 8.

```{r}
# save new test data frame, donʻt want to mess up airquality df
test_df <- airquality 

# writing a function to impute the mean by month
impute_mean <- function(x) {
  
  month_rows <- test_df[test_df$Month == x, ] 
  get_mean <- mean(month_rows$Solar.R, na.rm = TRUE) 
  month_rows[["Solar.R"]][is.na(month_rows[["Solar.R"]])] <- get_mean
  print(month_rows)
  
}

# iterate through each month on data set
may_avg <- impute_mean(x = 5)
june_avg <- impute_mean(x = 6)
july_avg <- impute_mean(x = 7)
august_avg <- impute_mean(x = 8)
september_avg <- impute_mean(x = 9)

# combine averages into concatinated data frame
impute_df <- rbind(may_avg, june_avg, july_avg, august_avg, september_avg)
impute_df

```

